---
title: "Assignment"
author: "Theo Burns"
date: 2/25/2026
format:
  html:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
---

## 

```{r}
library(tidyverse)

wdi <- read_csv("wdi.csv")
glimpse(wdi)
```

```{r}
summary(wdi$gdp_per_capita)
```

```{r}
#| echo: false
#| fig-cap: "Distribution of GDP per Capita across countries. Data source: World Bank World Development Indicators."
#| fig-label: fig-gdp-hist

ggplot(wdi, aes(x = gdp_per_capita)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of GDP per Capita",
    x = "GDP per Capita (USD)",
    y = "Number of Countries"
  )
```

As seen in @fig-gdp-hist, the distribution of GDP per capita is the most noteworthy. It is right skewed, with a large portion of countries in the 2 lowest bars. This indicates that most of the world is much poorer (on a per population basis) than the few richer countries. From @tbl-summary, it can be seen that the mean is much higher than the median, consistent with a right skewed set of data.

```{r}
#| echo: false
#| fig-cap: "Distribution of Inflation Rate across countries. Data source: World Bank World Development Indicators."
#| fig-label: fig-inflation-hist

ggplot(wdi, aes(x = inflation_rate)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Inflation Rate",
    x = "Inflation rate",
    y = "Number of Countries"
  )
```

```{r}
summary(wdi$inflation_rate)
```

As shown in @fig-inflation-hist, the distribution of inflation rate appears to be slightly right skewed, however it is much closer to a normal distribution than the GDP per capita data in @fig-gdp-hist.

```{r}
#| echo: false
#| fig-cap: "Distribution of Unemployment Rate across countries. Data source: World Bank World Development Indicators."
#| fig-label: fig-unemployment-hist

ggplot(wdi, aes(x = unemployment_rate)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Unemployment rate",
    x = "Unemployment rate",
    y = "Number of Countries"
  )
```

```{r}
summary(wdi$unemployment_rate)
```

@fig-unemployment-hist shows that the unemployment rate distribution is right skewed. This means that a majority of the unemployment data is relatively low, which is arguably an optimistic thing to observe. Coupled with the observation in @fig-gdp-hist of most countries having very low GDP per capita however, this potentially means that massive amounts of people are working for very poor wages.

```{r}
#| echo: false
#| fig-cap: "Relationship between GDP per capita and inflation rate across countries. Data source: World Bank World Development Indicators."
#| fig-label: fig-inflationrate-gdp

ggplot(wdi, aes(x = gdp_per_capita, y = inflation_rate)) +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  labs(
    title = "Inflation Rate vs GDP per Capita",
    x = "GDP per Capita (Log Scale, USD)",
    y = "Inflation Rate (%)"
  )
```

@fig-inflationrate-gdp appears to show that there is no observable relationship between GDP per Capita and inflation rate. Rising GDP per capita is known to lead to increasing inflation [@xiong2023], so this is an interesting result. Xong's finding, however, pertains to trends within countries rather than across countries, so the two findings are not necessarily conflicting.

```{r}
#| echo: false
#| fig-cap: "Top 10 countries by unemployment rate. Data source: World Bank World Development Indicators."
#| fig-label: fig-top10-unemployment

top10 <- wdi %>%
  filter(!is.na(unemployment_rate)) %>%
  arrange(desc(unemployment_rate)) %>%
  slice_head(n = 10)

ggplot(top10, aes(x = reorder(country, unemployment_rate), y = unemployment_rate)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Countries by Unemployment rate",
    x = "Country",
    y = "Unemployment Rate"
  )
```

As shown in @fig-top10-unemployment, 8 out of the top 10 countries with the highest unemployment rates are located in Africa. This is consistent with a map from the CIA World Factbook for 2015-2017, showing a concentration of nations with the worst employment rates in Africa [@factsmaps2018]. @tbl-summary provides key summary statistics for all three indicators.

```{r}
### Key Statistics for my 3 Indicators of Interest
```

```{r}
#| echo: false
#| tbl-cap: "Table 1. Summary statistics for selected World Development Indicators. Data source: World Bank World Development Indicators."
#| label: tbl-summary

library(dplyr)
library(tidyr)
library(knitr)

summary_tbl <- wdi %>%
  summarise(
    `GDP per Capita (USD)` = list(gdp_per_capita),
    `Life Expectancy (Years)` = list(life_expectancy),
    `Unemployment Rate (%)` = list(unemployment_rate)
  ) %>%
  pivot_longer(cols = everything(), names_to = "Indicator", values_to = "values") %>%
  mutate(
    N = sapply(values, function(x) sum(!is.na(x))),
    Mean = sapply(values, function(x) mean(x, na.rm = TRUE)),
    Median = sapply(values, function(x) median(x, na.rm = TRUE)),
    Min = sapply(values, function(x) min(x, na.rm = TRUE)),
    Max = sapply(values, function(x) max(x, na.rm = TRUE))
  ) %>%
  select(Indicator, N, Mean, Median, Min, Max) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

kable(summary_tbl)
```
