---
title: "World Development Indicators Analysis"
author: "Theo Burns"
date: 2/25/2026
format:
  revealjs:
    theme: league
    transition: zoom
---

```{r, echo=FALSE}
library(tidyverse)
wdi <- read_csv("wdi.csv")
```

## Introduction

-   This slideshow will be an analysis of the World Bank World Development Indicators dataset
-   The 3 indicators I have chosen to focus on are:
    -   GDP per Capita
    -   Inflation Rate
    -   Unemployment Rate

## GDP per Capita

```{r}
#| echo: false
#| fig-cap: "Distribution of GDP per Capita across countries. Data source: World Bank World Development Indicators."
#| fig-label: fig-gdp-hist

ggplot(wdi, aes(x = gdp_per_capita)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of GDP per Capita",
    x = "GDP per Capita (USD)",
    y = "Number of Countries"
  )
```

-   GDP per capita is right skewed

-   This indicates that most of the world is much poorer (on a per population basis) than the few wealthier countries

## Inflation Rate

```{r}
#| echo: false
#| fig-cap: "Distribution of Inflation Rate across countries. Data source: World Bank World Development Indicators."
#| fig-label: fig-inflation-hist

ggplot(wdi, aes(x = inflation_rate)) +
  geom_histogram(bins = 50) +
  labs(
    title = "Inflation Rate",
    x = "Inflation rate",
    y = "Number of Countries"
  )
```

-   The distribution of inflation appears to be slightly right skewed, but much more normal than that of GDP per capita

## Unemployment Rate

```{r}
#| echo: false
#| fig-cap: "Distribution of Unemployment Rate across countries. Data source: World Bank World Development Indicators."
#| fig-label: fig-unemployment-hist

ggplot(wdi, aes(x = unemployment_rate)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Unemployment rate",
    x = "Unemployment rate",
    y = "Number of Countries"
  )
```

-    The unemployment rate distribution is right skewed and shows that a majority of the unemployment data is relatively low

-   Coupled with the observation of most countries having very low GDP per capita however, this potentially means that massive amounts of people are working for very poor wages.

## GDP per Capita plotted against Inflation Rate

```{r}
#| echo: false
#| fig-cap: "Relationship between GDP per capita and life expectancy across countries. Data source: World Bank World Development Indicators."
#| fig-label: fig-inflationrate-gdp

ggplot(wdi, aes(x = gdp_per_capita, y = inflation_rate)) +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  labs(
    title = "Inflation Rate vs GDP per Capita",
    x = "GDP per Capita (Log Scale, USD)",
    y = "Inflation Rate (%)"
  )
```

-    There is no observable relationship between GDP per Capita and inflation rate.

## Top 10 Countries with the Highest Unemployment Rates

```{r}
#| echo: false
#| fig-cap: "Top 10 countries by unemployment rate. Data source: World Bank World Development Indicators."
#| fig-label: fig-top10-unemployment

top10 <- wdi %>%
  filter(!is.na(unemployment_rate)) %>%
  arrange(desc(unemployment_rate)) %>%
  slice_head(n = 10)

ggplot(top10, aes(x = reorder(country, unemployment_rate), y = unemployment_rate)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Countries by Unemployment rate",
    x = "Country",
    y = "Unemployment Rate"
  )
```

-   8 out of the top 10 countries with the highest unemployment rates are located in Africa.

-   This is consistent with a map from the CIA World Factbook for 2015-2017, showing a concentration of nations with the worst employment rates in Africa
